DATA 621 - Business Analytics and Data Mining

### 2020-10-09 HW3/HW3_Final.Rmd
ambers updates; model 1

```{r loadData, include=FALSE}
library(tidyverse)
library(summarytools)
library(corrplot)
library(knitr)

@@

## DATA EXPLORATION
### Data Description

@@

```{r sumstat}
summarystat = stby(data = df, INDICES = df$target, FUN = psych::describe)
kable(summarystat[[1]][-13,-c(1,7)], 
      caption = "Descriptive Statistics: Crime Rate > Median", 
      digit = 2L)
kable(summarystat[[2]][-13,-c(1,7)], 
      caption = "Descriptive Statistics: Crime Rate < Median", 
      digit = 2L)
```

### Data Distribution

@@

```{r density}
variable = names(df)
df.new = df
denplot = function(i){
  df.new$x = df.new[,variable[i]]
  ggplot(df.new, aes(x, fill = factor(df.new$target))) + 
    ggtitle(sprintf("%s", variable[i])) + 
    geom_density(alpha = 0.4, show.legend = FALSE) + 
    theme(legend.position = "none",
          plot.title = element_text(size = 10, hjust = 0.5),
          axis.title.y = element_blank(),
          axis.title.x = element_blank())
}
gridExtra::grid.arrange(denplot(1),denplot(2),denplot(3),denplot(4),
                        denplot(5),denplot(6),denplot(7),denplot(8),
                        denplot(9),denplot(10),denplot(11),denplot(12),
                        layout_matrix = rbind(c(0,1,2), c(3,4,5), c(6,7,8), c(9,10,11)))
```

@@

```{r boxplot}
df.new = reshape::melt(df, id.vars= "target") %>%
  dplyr::filter(variable != "chas") %>%
  mutate(target = as.factor(target))
ggplot(df.new, aes(x = variable, y = log(value))) + 
  geom_boxplot(color = "blue", fill = "blue", alpha = 0.2, 
               notch = TRUE, notchwidth = 0.8, 
               outlier.colour = "red", outlier.fill = "red", outlier.size = 3) + 
  stat_summary(fun.y = mean, color = "red", geom = "point", shape = 16, size = 2) + 
  coord_flip() + 
  labs(title = "Boxplot of Predictor Variables")

@@

### Data Correlation

@@

```{r correlation, fig.height=6, fig.width=6}
cor.df = function(df){
  df.m = as.matrix(df)
  n = ncol(df.m)
  p.mat = matrix(NA, n, n)
  diag(p.mat) = 0
  for (i in 1:(n - 1)){
    for (j in (i + 1):n){
      tmp = cor.test(df.m[, i], df.m[, j])
      p.mat[i, j] = p.mat[j, i] = tmp$p.value
    }
  }
  colnames(p.mat) = rownames(p.mat) = colnames(df.m)
  p.mat
}
p.mat = cor.df(df)
M = cor(df)
col = colorRampPalette(c("#BB4444", "#EE9988", "#FFFFFF", "#77AADD", "#4477AA"))
corrplot(round(M,1), method = "color", col = col(200),  
         type = "lower", order = "hclust", 
         addCoef.col = "black",
         tl.col = "black", tl.srt = 45,
         p.mat = p.mat, sig.level = 0.01, insig = "blank", 
         diag = FALSE)
```

## APPENDIX
The code chunks below represent the R code called in order during the analysis. They are reproduced in the appendix for review and comment.
```{r appendix, include=FALSE}
knitr::opts_chunk$set(echo=TRUE, eval=FALSE)
```
```{r loadData}
```
```{r sumstat}
```
```{r density}
```
```{r boxplot}
```
```{r correlation}
```

>>>>

```{r loadData, include=FALSE}
library(tidyverse)
library(summarytools)
library(corrplot)
library(knitr)
library(rsample) # model 1 libraries
library(caret)
library(e1071)

set.seed(9450)

@@

# DATA EXPLORATION
## Data Description

@@

```{r sumstat}
# DATA EXPLORATION 
summarystat = stby(data = df, INDICES = df$target, FUN = psych::describe)
kable(summarystat[[1]][-13,-c(1,7)], 
      caption = "Descriptive Statistics: Crime Rate > Median", 
      digit = 2L)
kable(summarystat[[2]][-13,-c(1,7)], 
      caption = "Descriptive Statistics: Crime Rate < Median", 
      digit = 2L)
```

## Data Distribution

@@

```{r density}
# density plots
variable = names(df)
df.new = df
denplot = function(i){
  df.new$x = df.new[,variable[i]]
  ggplot(df.new, aes(x, fill = factor(target))) + 
    ggtitle(sprintf("%s", variable[i])) + 
    geom_density(alpha = 0.4, show.legend = FALSE) + 
    theme(legend.position = "none",
          plot.title = element_text(size = 10, hjust = 0.5),
          axis.title.y = element_blank(),
          axis.title.x = element_blank())
}
gridExtra::grid.arrange(denplot(1),denplot(2),denplot(3),denplot(4),
                        denplot(5),denplot(6),denplot(7),denplot(8),
                        denplot(9),denplot(10),denplot(11),denplot(12),
                        layout_matrix = rbind(c(0,1,2), c(3,4,5), c(6,7,8), c(9,10,11)))
```

@@

```{r boxplot}
# boxplots 
df.new = reshape::melt(df, id.vars= "target") %>%
  dplyr::filter(variable != "chas") %>%
  mutate(target = as.factor(target))
ggplot(df.new, aes(x = variable, y = log(value))) + 
  geom_boxplot(color = "blue", fill = "blue", alpha = 0.2, 
               notch = TRUE, notchwidth = 0.8, 
               outlier.colour = "red", outlier.fill = "red", outlier.size = 3) + 
  stat_summary(fun.y = mean, color = "red", geom = "point", shape = 16, size = 2) + 
  coord_flip() + 
  labs(title = "Boxplot of Predictor Variables")

@@

## Data Correlation

@@

```{r correlation, fig.height=6, fig.width=6}
# correlations
cor.df = function(df){
  df.m = as.matrix(df)
  n = ncol(df.m)
  p.mat = matrix(NA, n, n)
  diag(p.mat) = 0
  for (i in 1:(n - 1)){
    for (j in (i + 1):n){
      tmp = cor.test(df.m[, i], df.m[, j])
      p.mat[i, j] = p.mat[j, i] = tmp$p.value
    }
  }
  colnames(p.mat) = rownames(p.mat) = colnames(df.m)
  p.mat
}
p.mat = cor.df(df)
M = cor(df)
col = colorRampPalette(c("#BB4444", "#EE9988", "#FFFFFF", "#77AADD", "#4477AA"))
corrplot(round(M,1), method = "color", col = col(200),  
         type = "lower", order = "hclust", 
         addCoef.col = "black",
         tl.col = "black", tl.srt = 45,
         p.mat = p.mat, sig.level = 0.01, insig = "blank", 
         diag = FALSE)
```

# DATA PREPARATION

# BUILD MODELS

First, we will split the training data into two sets. This will allow us to perform a cross validation scheme on the models to tune parameters for optimum performance.

```{r tts}

# BUILD MODELS
# train-test split
crime_df<-
  df %>% 
  mutate(target = as.factor(target)) %>% 
  mutate_if(is.integer, as.numeric)

data_split <- initial_split(crime_df, 
                            strata = target, 
                            prop = 0.8)
train_df<- training(data_split)
test_df<- testing(data_split)
```

## Model 1
### Base Model
We will start with a simple logistic regression model to serve as a baseline. This includes
all variables in the dataset.

```{r baseModel}

baseModel <- glm(target~., family = binomial, data = train_df)

knitr::kable(summary(baseModel)$coefficients, digits = 3L,
             caption = 'Base Model Logistic Regression Output')

```
We can immediately see that a few variables *exceed* the 0.05 p-value threshold
for significance.

### Enhanced Model
We will use **backwards stepwise regression** to remove features that are not statistically significant in predicting the target. The result is a model that includes the following features: `zn`, `nox`, `age`, `dis`, `rad`, `tax`, `ptratio`, `lstat`, and `medv`.

### Coefficient Discussion
It is important to note that the coefficients from the model are predicting whether or not the target variable is a **1** (the crime rate is *above* the median value). Additionally, it must be noted that the numeric coefficients are relative to the range of values that the variable encompasses. What this means is that it's possible to have a coefficient that seems small when we look at the absolute magnitude, but that actually has a very strong effect when applied to the data. 
```{r model1_coeff}

# 10-fold cross validation
train_control <- trainControl(method = "cv", number = 10)

# train the model on training set
model1 <- train(target ~ zn+nox+age+dis+rad+tax+ptratio+lstat+medv,
               data = train_df,
               trControl = train_control,
               method = "glm",
               family=binomial())

knitr::kable(summary(model1)$coefficients, digits = 3L,
             caption = 'Backwards Model Regression Output')

```
Let's take a look at the *signs* of the coefficients. We can see that the coefficients for the variables `zn` and `tax` are negative. This is indicative of an inverse relationship; ie, the higher these values, the less likely the crime rate is above the median. The relationship for the zone variable aligns with the findings from our initial data exploration. However, the relationship for the tax variable does not; in fact, we saw that there was a positive correlation between the two (as the tax rate increases, so does the probability that the crime rate will be above the median).

The rest of the variables have positive coefficients. Many of these are expected. For example, in our exploratory data analysis, we noticed that the `nox` variable (nitrogen oxides concentration (parts per 10 million)) has the greatest positive correlation with the target variable. This is surprising; we hadn't thought that the nitrogen oxide concentration would have as large of an impact on the crime rate, but some quick googling shows us that there may actually be a relationship. ( [https://www.sciencedaily.com/releases/2019/10/191003114007.htm](Source) )

# SELECT MODELS
## Criteria

## Performance

### Model 1
```{r model1_perf}

# SELECT MODELS
# Model 1 Performance
model1_preds <- predict(model1, test_df, type = "raw")
model1_probs <- predict(model1, test_df, type = "prob")
colnames(model1_probs) <- c('pred0', 'pred1')

model1_results <- test_df %>%
  bind_cols(pred = model1_preds, model1_probs)

####### metrics
m1_roc<- yardstick::roc_auc(
  model1_results,
  truth = target,
  pred0 # select the prob class that corresponds to first level of target
)

m1_acc<- yardstick::accuracy(
  model1_results, 
  truth = target, 
  estimate = pred
)

m1_recall<- yardstick::recall(
  model1_results,
  truth = target, 
  estimate = pred
)

m1_precise<- yardstick::precision(
  model1_results,
  truth = target, 
  estimate = pred
)

m1_metrics_df<-
  bind_rows(m1_roc, m1_acc, m1_recall, m1_precise)

m1_metrics_df

```

# APPENDIX
The code chunks below represent the R code called in order during the analysis. They are reproduced in the appendix for review and comment.
```{r appendix, include=FALSE}
knitr::opts_chunk$set(echo=TRUE, eval=FALSE)
```
```{r loadData}
```
```{r sumstat}
```
```{r density}
```
```{r boxplot}
```
```{r correlation}
```
